Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-23 paper/reviewer_response_PNAS.tex

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

\response{
This is an excellent question.
We now a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
Because sequences from unpassaged isolates have only become common relatively recently, we could perform this analysis only for the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
The new Figure 5 shows that the correlation between our measurements and natural evolution is actually \emph{stronger} when we look only at unpassaged isolates.
This makes sense---lab passage mutations will not do well in nature, but will be scored favorably in our experiments.
Therefore, it makes sense that our approach works \emph{even better} when the analysis is restricted to unpassaged isolates.
We thank the reviewer for this suggestion, which has strengthened our work.
}

5) It is unclear how the ``rescaling'' is performed and what it exactly does. The authors need to described and justify this better and should also provide the non-rescaled data in a ``sequence logo'' figure for comparison.

\response{
The re-scaling of amino-acid preferences by a stringency parameter is a well-defined mathematical procedure that is explained in detail in Hilton et al (\textit{PeerJ}, 5:e3657, 2017), which is cited in the manuscript.
This re-scaling procedure has now been used fairly widely in the literature. 
Importantly, re-scaling does \emph{not} affect any of the evolutionary analyses performed in this paper since it just corresponds to a constant that multiplies all mutational effects (e.g., a change in units on the x-axis of Figure~5).
Specifically, if the re-scaling parameter is $\beta$, then the mutational effects calculated using Equation~1 in the manuscript are just multiplied by $\beta$ after re-scaling.
Therefore, for analyses of mutational effects, the re-scaling simply corresponds to a change in units (e.g., comparable to showing an antibody titration in ng/mL versus $\mu$M), and therefore has no effect on the conclusions.
We have added the following text in the Methods explaining this fact:

\begin{quote}
\textsl{The amino-acid preferences were re-scaled by the stringency parameter using the approach described in Hilton et al (2017).
Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
The reason is that all of these conclusions use the effects of mutations as calculated using Equation~1, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.}
\end{quote}

Due to space constraints and the size of the logo plots, we cannot fit another main figure with the preferences prior to re-scaling, but such an image is now included in the Jupyter notebook in the manuscript's GitHub repository at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
The unscaled amino-acid preferences for each replicate are also provided in numerical form in Dataset S3.
}

### 2018-06-23 paper/manuscript.tex
comment 5 reviewer 1

\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.
The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.

@@

\subsection*{Quantification of mutational effects}
We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
The effect of mutating site $r$ from amino-acid $a_1$ to $a_2$ was calculated as
\begin{equation}
\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
\end{equation}
where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).

>>>>

\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.

The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
The reason is that all of these conclusions use the effects of mutations as calculated using Equation~\ref{eq:muteffect}, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.

@@

\subsection*{Quantification of mutational effects}
We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
The effect of mutating site $r$ from amino-acid $a_1$ to $a_2$ was calculated as
\begin{equation}
\label{eq:muteffect}
\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
\end{equation}
where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
